Impact of grids and dynamical cores in CESM2.2 on the surface mass balance of the Greenland Ice Sheet

### 2022-04-28 arcticgrids.tex
new grid figures, marcus edits

Six different configurations, a mixture of grids and atmospheric dynamical cores available in the Community Earth System Model, version 2.2 (CESM2.2), are evaluated for their skill in representing the climate of the Arctic and the surface mass balance of the Greenland Ice Sheet (GrIS). The more conventional $1^{\circ}-2^{\circ}$ uniform resolution grids systematically overestimate both accumulation and ablation over the GrIS. Of these conventional grids, the latitude-longitude grids outperform the quasi-uniform unstructured grids owing to their greater degrees of freedom in representing the GrIS. Two Arctic-refined meshes, with $1/4^{\circ}$ and $1/8^{\circ}$ refinement over Greenland, are documented as newly supported configurations in CESM2.2. The Arctic meshes substantially improve the simulated clouds and precipitation rates in the Arctic. Over Greenland, these meshes skillfully represent accumulation and ablation processes, leading to a more realistic GrIS surface mass balance. As CESM is in the process of transitioning away from conventional latitude-longitude grids, these new Arctic refined meshes improve the representation of polar processes in CESM by recovering resolution lost in the transition to quasi-uniform grids.

@@

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high latitudes, which are among the most sensitive regions on Earth to global and environmental change. GCMs differ vastly in their numerical treatment of polar regions because of the so-called \textit{pole-problem} \cite{W2007JMSJ}. The pole problem refers to numerical instability arising from the convergence of meridian lines into polar singularities on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a, hereafter referred to as \textit{lat-lon} grids). Depending on the numerics, methods exist to suppress this instability, and lat-lon grids may be advantageous for polar processes by representing structures with finer resolution than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of lat-lon grids in polar regions may be lost. In this study we evaluate a number of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 \cite<CESM2.2;>{CESM2}, including new variable-resolution grids, to understand their impacts on the simulated Arctic climate. We focus specifically on the climate and surface mass balance of the Greenland Ice Sheet.

In the 1970s, the pole problem was largely defeated through the adoption of efficient spectral transform methods in GCMs \cite<see>[and references therein]{W2007JMSJ}.
These methods transform grid point fields into a global, isotropic representation in wave space, where linear operators (e.g., horizontal derivatives) in the (truncated) equation set can be solved exactly. While spectral transform methods are still used today, local numerical methods have become desirable for their ability to run efficiently on massively parallel systems. The pole problem has thus re-emerged in contemporary climate models that use lat-lon grids, and some combination of reduced grids (modified lat-lon grids, with cells elongated in the longitudinal direction over the polar regions) and polar filters are necessary to ameliorate this instability \cite{JW2010LNCSE}. Polar filters subdue the growth of unstable modes by applying additional damping to the solution over polar regions. This damping reduces the effective resolution in polar regions such that the resolved scales are \textit{approximately} the same everywhere on the grid. We emphasize \textit{approximately} since it's at least conceivable that marginal increases in effective resolution occur over polar region in lat-lon grids, despite polar filtering, since resolved waves can be represented with more grid points than at lower latitudes.

Dycores built on lat-lon grids have some advantages over unstructured grids. Lat-lon coordinate lines are orthogonal, and aligned with zonally symmetric circulations that characterize many large-scale features of Earth's atmosphere. \citeA{LJTN2010JAMES} has experimented with rotating lat-lon models such that their coordinate lines no longer align with an idealized, zonally balanced circulation. For the finite-volume lat-lon dycore considered in this paper (hereafter \textit{FV}), numerical errors were shown to be largest when the polar singularity is rotated into the baroclinic zone ($45^{\circ}$N latitude), generating spurious wave growth much earlier on in the simulation than for other rotation angles. This illustrates the advantages of coordinate surfaces aligned with latitude bands, albeit an extreme example where the polar singularity and the polar filter are also contributing to the spurious wave growth. The unstructured grids all generate spurious baroclinic waves earlier on in the simulations than the (unrotated) lat-lon models, although the unstructured model considered in this paper, the spectral-element dycore (hereafter \textit{SE}), holds a balanced zonal flow without wave growth appreciably longer than the rotated FV experiments, regardless of how the SE grid is rotated \cite{LJTN2010JAMES}.

The polar filter in the FV model impedes efficiency at large processor (CPU) counts because it requires a spectral transform, which have large communication overhead \cite{ST1995GEOS,DetAl2012IJHPCA}. Unstructured grids support quasi-uniform grid spacing globally, and there is no pole-problem (e.g., Figure~\ref{fig:uni-grids}c). Unstructured grids are becoming increasingly common due to their improved performance on massively parallel systems and lack of constraints on grid structure \cite{PL2007JCP,DetAl2012IJHPCA,WETAL2013GMD}. This grid flexibility allows for variable-resolution grids (e.g., Figure~\ref{fig:vr-grids}; hereafter abbreviated as \textit{VR}), sometimes referred to as regional grid refinement. In principle, grid refinement over polar regions can make up for any loss of resolution in transitioning away from lat-lon grids (e.g., Figure~\ref{fig:vr-grids}).  However, refinement comes at the cost of a smaller CFL-limited time step in the refined region. (The CFL-condition --- short for Courant–Friedrichs–Lewy condition --- is a necessary condition for numerical stability when using discrete data in time and space.)

It is important to emphasize that the pole-problem is a feature of the dycore. Polar filters do not directly interfere with the physical parameterizations, nor do they have any bearing on the surface models; e.g., the land model can take full advantage of the greater number of grid cells in polar regions on lat-lon grids. This is particularly relevant for the surface mass balance of the Greenland Ice Sheet (\textit{SMB}; the integrated sum of precipitation and runoff), which relies on hydrological processes handled by the land model.

@@

Regional climate models (RCMs) are commonly relied upon to provide more accurate SMB estimates. The limited area domain used by RCMs permits the use of high resolution grids able to resolve the SMB processes, and can skillfully simulate the GrIS SMB \cite{BETAL2004JGR,RETAL2012TC,VETAL2012TC,FETAL2013TC,METAL2017LTS,NETAL2018TC}. However, unlike GCMs, RCMs are not a freely evolving system and boundary conditions must be prescribed at the lateral boundaries. The inability of the RCM solution to influence larger-scale dynamics outside the RCM domain (due to the prescribed boundary conditions) severely limits this approach from properly representing the role of the GrIS in the climate system. In addition, the boundary conditions are derived from a separate host model, which introduces inconsistencies due to differences in model design between the host model and the RCM.

@@

The SE dycore has been an available option since CESM, version 1, but has been under active development ever since. This includes the switch to a dry-mass vertical coordinate \cite{LetAl2018JAMES} and incorporation of an accelerated multi-tracer transport scheme \cite{LTOUNGK2017MWR}, made available in CESM2. This paper documents several additional enhancements to the SE dycore as part of the release of CESM2.2. These include three new VR configurations (Figure~\ref{fig:vr-grids}), two Arctic meshes and a Contiguous United-States mesh ($\tt CONUS$; featured in \citeA{PETAL2020BAMS}). While there are dozens of published studies using VR in CESM \cite<e.g., >{ZetAl2014JC,RHUZ2016JAMC,GetAl2017JAMES,BETAL2019TAC,BETAL2021IJC}, these studies either used development code or collaborated closely with model developers. CESM2.2 is the first code release that contains out of the box VR functionality in CESM.

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{cc}
         \includegraphics[width=60mm]{figs/grid-f19.pdf}&
         \includegraphics[width=60mm]{figs/grid-f09.pdf}\\
         \includegraphics[width=60mm]{figs/grid-ne30pg2.pdf}&
         \includegraphics[width=60mm]{figs/grid-ne30pg3.pdf} \\
\end{tabular}
\end{center}
\caption{Computational grids for the uniform $1^{\circ}-2^{\circ}$ grids in this study. Grids names after Table~\ref{tbl:table1}, (a) $\tt f19$, (b) $\tt f09$, (c) $\tt ne30pg2$ and (d) $\tt ne30pg3$.}
\label{fig:uni-grids}
\end{figure}

\begin{figure}[t]
\begin{center}
\begin{tabular}{cccc}
         \includegraphics[width=60mm]{figs/grid-ARCTIC.pdf}&
         \includegraphics[width=60mm]{figs/grid-ARCTICGRIS.pdf} \\
\end{tabular}
\end{center}
\caption{Spectral-element grid for the VR grids in this study, (a)$\tt Arctic$, (b) $\tt Arctic-GrIS$ and (a) $\tt CONUS$. Note that this is not the computational grid; each element has $3\times3$ independent grid points. {\color{red}{add CONUS grid}}}
\label{fig:vr-grids}
\end{figure}

@@

The FV dycore is a hydrostatic model that integrates the equations of motion using a finite-volume discretization on a spherical lat-lon grid \cite{LR1997QJR}. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to an Eulerian reference grid in the vertical \cite{L2004MWR}, using a hybrid-pressure vertical coordinate. Hyperviscous damping is applied to the divergent modes, while Laplacian damping is applied to momentum in the top few layers, referred to as a \textit{sponge layer} \cite{L2011IJHPC} [phl: as far as I know the Laplacian-like damping is not invoked in lower resolution setups - was implemented for 1/4 applications where the polar night jet became excessively strong]. A polar filter damps computational instability due to the convergence of meridians, permitting a more practical time step. It takes the form of a Fourier filter in the zonal direction, with the damping coefficients increasing monotonically in the meridional direction \cite{ST1995GEOS}. The form of the filter is designed to slow down the propagation of large zonal wave-numbers to satisfy the CFL condition of the shortest resolved wave at some reference latitude.

@@

The SE dycore supports regional grid refinement via its VR configuration, requiring two enhancements over uniform-resolution setups. (1) As the numerical viscosity increases with resolution, explicit hyperviscosity relaxes according to the local element size, reducing in strength  by an order of magnitude per halving of the grid spacing. A tensor-hyperviscosity formulation is used \cite{GetAl2014GMD}, which adjusts the coefficients in two orthogonal directions to more accurately target highly distorted quadrilateral elements. (2) The topography boundary conditions are smoothed in a way that does not excite grid scale modes, and so the NCAR topography software \cite{gmdd-8-4623-2015} has been modified to scale the smoothing radius by the local element size, resulting in rougher topography in the refinement zone.

For SE grids with quasi-uniform grid spacing, The SE tracer transport scheme is replaced with the Conservative Semi-Lagrangian Multi-tracer transport scheme (CSLAM) \cite<>{LTOUNGK2017MWR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a separate grid from the spectral-element dynamics, dividing each element into $3\times3$ control volumes with quasi-equal area. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the default SE dycore in which the physics are evaluated Gauss-Lobatto-Legendre points \cite{HL2018MWR}. {\color{purple}{Rene/Andrew - maybe elaborate further on the advantages of CSLAM?}}

@@

All grids and dycores in this study use 32 levels in the vertical, with a model top of about 1~hPa or 40~km. However, note that any grid or dycore can in principle be run with a higher model top or finer vertical resolution.

@@

Initial simulations with the $\tt ne30pg3$ SE grid produced weaker shortwave cloud forcing relative to the tuned finite-volume dycore. The SE dycore in CESM2.2 therefore has two CLUBB parameter changes to provide more realistic cloud forcing and top-of-atmosphere radiation balance. We reduced the width of the sub-grid distribution of vertical velocity ($\tt clubb\_gamma=0.308\rightarrow0.270$) and also reduced the strength of the damping for horizontal component of turbulent energy ($\tt clubb\_c14=2.2\rightarrow1.6$) to increase cloudiness. For a thorough explanation of how CLUBB parameters impact the simulated climate, see \cite{GETAL2015JAMES}.

\subsection{Modeled SMB}

All grids and dycores simulate the GrIS Surface Mass Balance (SMB). The SMB is the sum of mass accumulation of precipitation and mass loss from ablation. Ablation is the sum of evaporation/sublimation and liquid runoff, with runoff being a combination of liquid precipitation and snow and ice melt. Not all liquid precipitation or snow/ice melt runs off the ice sheet; this water can penetrate pore spaces in the snowpack/firn layer and freeze, increasing the ice mass. These processes are represented by different CESM components, but it is the Community Land Model, version 5 \cite<CLM;>{CLM5}, that aggregates these processes and computes the SMB.

CLM runs on the same grid as the atmosphere, using a downscaling technique to account for sub-grid variability in SMB. In short, the ice sheet patch in a CLM grid cell is subdivided into 10 elevation classes (ECs), each with a distinct surface energy balance and SMB. The area fraction of each EC is derived from a high-resolution GrIS elevation dataset. The near-surface air temperature, humidity, and air density are calculated for each EC using an assumed lapse rate and the elevation difference from the grid-cell mean. Precipitation from CAM is repartitioned into solid or liquid based on the surface temperature of the EC; precipitation falls as snow for temperatures between $T < -2^\circ$~C, as rain for $T > 0^\circ$~C,  and as a linear combination of rain and snow for temperatures between $-2^\circ$~C and $0^\circ$~C.
Snow accumulation in each EC is limited to a depth of 10 m liquid water equivalent. Any snow above the 10-m cap is routed to the ocean as solid runoff.  Refreezing of liquid water within the snowpack is an additional source of ice.  Integrating over all ECs, weighting by the area fractions, provides a more accurate SMB than would be found using the grid-cell mean elevation. For a more detailed description of how the SMB is computed in CESM, we refer the reader to \citeA{CISM1,SETAL2019TC,KETAL2020JAMES,Muntjewerf_EA_2021_coupling}. {\color{orange}{Marcus - mention you're not running CISM.}}

Changes in ice depth, but not snow depth, count toward the SMB.  That is, snow accumulation above the 10-m cap contributes a positive SMB, and surface ice melting (after melting of the overlying snow) yields a negative SMB.  Since snow in the accumulation zone must reach the cap to simulate a positive SMB, the snow depths on the VR grids were spun up by forcing CLM in standalone mode, cycling over a 20-year $\tt Arctic$ $FHIST$ run for about 500 years. The uniform-resolution grids are initialized with the SMB from an existing $\tt f09$ spun-up initial condition.

@@

We seek to integrate SMB components over a GrIS ice mask and to diagnose their contributions to the GrIS mass budget. However, the ice masks vary across the grids, especially in comparison to the RACMO3.2 ice mask, whose total area is about $3\%$ less than that of the reference dataset (Figure~\ref{fig:grisdx}). The area errors in Figure~\ref{fig:grisdx} may not seem large, but even $1-2\%$ area differences can lead to large differences in integrated SMB \cite{HETAL2022TC}. CLM's dataset creation tool generates the model ice mask by mapping a high-resolution dataset to the target grid using the Earth System Modeling Framework (ESMF) first-order conservative remapping algorithm \cite{ESMF}. The figure suggests that the mapping errors are less than $1.5\%$ across the CESM2 grids.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_grisres.pdf}
\end{center}
\caption{The spatial properties of the GrIS as represented by different grids in this study. (Left) GrIS area error, computed as the relative differences from a 4km dataset used to create the CESM ice masks, (right) approximate average grid spacing over GrIS.}
\label{fig:grisdx}
\end{figure}

We have taken a common-ice-mask approach by mapping all model fields to the lowest-resolution grids, i.e., the $\tt f19$ and $\tt ne30pg2$ grids, and integrating over these low-resolution ice masks. The use of low-resolution common ice masks is a conservative decision and is largely because we seek to use first-order remapping algorithms to map fields to the common ice mask, which is not reliable in the downscaling direction, i.e., mapping to a higher-resolution grid than the source grid. We use two remapping algorithms: ESMF first-order conservative and the TempestRemap \cite{TempestRemap} high-order monotone algorithm. Since mapping errors are sensitive to grid type, we evaluate all quantities on both common ice masks, the $\tt f19$ and $\tt ne30pg2$ masks. Thus, we evaluate an integrated quantity on a given grid up to four times to estimate the uncertainty due to differences in grid type and remapping algorithms.

The SMB is expressed in a form that is agnostic of water phase, a total water mass balance, to facilitate comparisons across different grids with different ice masks and to increase consistency with the variables available in the RACMO datasets. The SMB for total water can be expressed as:
\begin{equation}
SMB = precipitation + runoff + evaporation/sublimation, \label{eq:SMB}
\end{equation}
where all terms have consistent sign conventions (positive values contribute mass, negative values reduce mass). The precipitation source term refers to the combined solid and liquid precipitation, runoff refers to the liquid water sink, and evaporation/sublimation is the vapor sink. Since the runoff term aggregates many processes, we isolate the melting contribution by also tracking the combined snow plus ice melt. Note that this SMB expression is different from the internally computed SMB in CLM \cite{KETAL2020JAMES}.

@@

Note that we are mapping to low-resolution grids that have larger GrIS areas (Figure~\ref{fig:grisdx}). Since the components of equation~\ref{eq:SMB} are not confined to the ice mask, method~\ref{lbl:label1} reconstructs the SMB over the portion of the common ice mask that is outside the ice mask on the source grid. While this may be a an acceptable way to reconstruct the mass source terms over different ice masks, ice melt is zero outside the source ice mask, and so method~\ref{lbl:label1} will underestimate the mass sink term. This underestimation is systematic in method~\ref{lbl:label2}; all variables are exclusive to the ice mask. Method~\ref{lbl:label2} is more dissipative, since mapping to the low-resolution grid will average a field of non-zero values over the ice mask with a field of zeros outside the ice mask. However, patch-level values for processes exclusive to the ice mask (e.g., ice melt) will on average have larger magnitudes than the the grid-mean quantities used in method~\ref{lbl:label1}. %The dilution of the these patch level fields caused by mapping to the common ice mask are then inflated due to integrating over the larger ice mask.

>>>>

Six different configurations, a mixture of grids and atmospheric dynamical cores available in the Community Earth System Model, version 2.2 (CESM2.2), are evaluated for their skill in representing the climate of the Arctic and the surface mass balance of the Greenland Ice Sheet (GrIS). The conventional $1^{\circ}-2^{\circ}$ uniform resolution grids systematically overestimate both accumulation and ablation over the GrIS. Of these conventional grids, the latitude-longitude grids outperform the quasi-uniform unstructured grids owing to their higher degrees of freedom in representing the GrIS. Two Arctic-refined meshes, with $1/4^{\circ}$ and $1/8^{\circ}$ refinement over Greenland, are documented as newly supported configurations in CESM2.2. The Arctic meshes substantially improve the simulated clouds and precipitation rates in the Arctic. Over Greenland, these meshes skillfully represent accumulation and ablation processes, leading to a more realistic GrIS surface mass balance. As CESM is in the process of transitioning away from conventional latitude-longitude grids, these new Arctic refined meshes improve the representation of polar processes in CESM by recovering resolution lost in the transition to quasi-uniform grids.

@@

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high latitudes, which are among the most sensitive regions on Earth to global and environmental change. GCMs differ vastly in their numerical treatment of polar regions because of the so-called \textit{pole-problem} \cite{W2007JMSJ}. The pole problem refers to numerical instability arising from the convergence of meridian lines into polar singularities on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a, hereafter referred to as \textit{lat-lon} grids). Depending on the numerics, methods exist to suppress this instability, and lat-lon grids may be advantageous for polar processes by representing structures with finer resolution than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of lat-lon grids in polar regions may be lost. In this study, we evaluate a number of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 \cite<CESM2.2;>{CESM2}, including new variable-resolution grids, to understand their impacts on the simulated Arctic climate. We focus specifically on the climate and surface mass balance of the Greenland Ice Sheet.

In the 1970s, the pole problem was largely defeated through the adoption of efficient spectral transform methods in GCMs \cite<see>[and references therein]{W2007JMSJ}.
These methods transform grid point fields into a global, isotropic representation in wave space, where linear operators (e.g., horizontal derivatives) in the (truncated) equation set can be solved exactly. While spectral transform methods are still used today, local numerical methods have become desirable for their ability to run efficiently on massively parallel systems. The pole problem has thus re-emerged in contemporary climate models that use lat-lon grids, and some combination of reduced grids (modified lat-lon grids, with cells elongated in the longitudinal direction over the polar regions) and polar filters are necessary to ameliorate this numerical instability \cite{JW2010LNCSE}. Polar filters subdue the growth of unstable computational modes by applying additional damping to the numerical solution over polar regions. This damping reduces the effective resolution in polar regions such that the resolved scales are \textit{approximately} the same everywhere on the grid. We emphasize \textit{approximately}, since it is at least conceivable that marginal increases in effective resolution occur over polar region in lat-lon grids, despite polar filtering, since resolved waves can be represented with more grid points than at lower latitudes.

Dycores built on lat-lon grids have some advantages over unstructured grids. Lat-lon coordinate lines are orthogonal, and aligned with zonally symmetric circulations that characterize many large-scale features of Earth's atmosphere. \citeA{LJTN2010JAMES} has experimented with rotating lat-lon models such that their coordinate lines no longer align with an idealized, zonally balanced circulation. For the finite-volume lat-lon dycore considered in this paper (hereafter \textit{FV}), numerical errors were shown to be largest when the polar singularity is rotated into the baroclinic zone ($45^{\circ}$N latitude), generating spurious wave growth much earlier in the simulation than for other rotation angles. This illustrates the advantages of coordinate surfaces aligned with latitude bands, albeit an extreme example where the polar singularity and the polar filter are also contributing to the spurious wave growth. The unstructured grids all generate spurious baroclinic waves earlier on in the simulations than the (unrotated) lat-lon models, although the unstructured model considered in this paper, the spectral-element dycore (hereafter \textit{SE}), holds a balanced zonal flow without wave growth appreciably longer than the rotated FV experiments, regardless of how the SE grid is rotated \cite{LJTN2010JAMES}. {\color{purple}{Marcus - this sentence is very complicating.}}

The polar filter in the FV model impedes efficiency at large processor (CPU) counts because it requires a spectral transform, which have large communication overhead \cite{ST1995GEOS,DetAl2012IJHPCA}. Unstructured grids support quasi-uniform grid spacing globally, and there is no pole-problem (e.g., Figure~\ref{fig:uni-grids}c). Conversely, unstructured grids are becoming increasingly common due to their improved performance on massively parallel systems and lack of constraints on grid structure \cite{PL2007JCP,DetAl2012IJHPCA,WETAL2013GMD}. This grid flexibility allows for the adoption of nested variable-resolution grids (e.g., Figure~\ref{fig:vr-grids}; hereafter abbreviated as \textit{VR}), sometimes referred to as regional grid refinement. In principle, grid refinement over polar regions can make up for any loss of resolution in transitioning away from lat-lon grids (e.g., Figure~\ref{fig:vr-grids}).  However, local grid refinement comes at the cost of a smaller CFL-limited time step in the refined region; the CFL-condition --- short for Courant–Friedrichs–Lewy condition --- is a necessary condition for numerical stability when using discrete data in time and space.

It is important to emphasize that the pole-problem is a distinctive feature of the dycore in atmospheric models. Polar filters do not directly interfere with the physical parameterizations, nor do they have any bearing on the surface models; e.g., the land model can take full advantage of the greater number of grid cells in polar regions on lat-lon grids. This is particularly relevant for the surface mass balance of the Greenland Ice Sheet (\textit{SMB}; the integrated sum of precipitation and runoff), which relies on hydrological processes represented in the land model.

@@

Regional climate models (RCMs) are commonly relied upon to provide more accurate SMB estimates. The limited area domain used by RCMs permits the use of high resolution grids, capable of resolving SMB processes, that can skillfully simulate the GrIS SMB \cite{BETAL2004JGR,RETAL2012TC,VETAL2012TC,FETAL2013TC,METAL2017LTS,NETAL2018TC}. However, unlike GCMs, RCMs are not a freely evolving system and the atmospheric state must be prescribed at the lateral boundaries of the model domain. The inability of the RCM solution to influence larger-scale dynamics outside the RCM domain (due to the prescribed boundary conditions) severely limits this approach from properly representing the role of the GrIS in the climate system. In addition, the boundary conditions are derived from a separate host model, which introduces inconsistencies due to differences in model design between the host model and the RCM.

@@

The SE dycore has been included in the model since CESM, version 1, but has been under active development ever since. This includes the switch to a dry-mass vertical coordinate \cite{LetAl2018JAMES} and incorporation of an accelerated multi-tracer transport scheme \cite{LTOUNGK2017MWR}, made available in CESM2. This paper documents several additional enhancements to the SE dycore as part of the release of CESM2.2. These include three new VR configurations (Figure~\ref{fig:vr-grids}), two Arctic meshes and a Contiguous United-States mesh ($\tt CONUS$; featured in \citeA{PETAL2020BAMS}). While there are dozens of published studies using VR in CESM \cite<e.g., >{ZetAl2014JC,RHUZ2016JAMC,GetAl2017JAMES,BETAL2019TAC,BETAL2021IJC}, these studies either used development code or collaborated closely with model developers. CESM2.2 is the first code release that contains out of the box VR functionality in CESM.

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{cc}
         \includegraphics[width=50mm]{figs/grid-f19.pdf}&
         \includegraphics[width=50mm]{figs/grid-f09.pdf}\\
         \includegraphics[width=50mm]{figs/grid-ne30pg2.pdf}&
         \includegraphics[width=50mm]{figs/grid-ne30pg3.pdf} \\
\end{tabular}
\end{center}
\caption{Computational grids for the uniform $1^{\circ}-2^{\circ}$ grids in this study. Grids names after Table~\ref{tbl:table1}, (a) $\tt f19$, (b) $\tt f09$, (c) $\tt ne30pg2$ and (d) $\tt ne30pg3$.}
\label{fig:uni-grids}
\end{figure}

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
         \includegraphics[width=40mm]{figs/grid-ARCTIC.pdf}&
         \includegraphics[width=40mm]{figs/grid-ARCTICGRIS.pdf}&
         \includegraphics[width=40mm]{figs/grid-CONUS.pdf} \\
\end{tabular}
\end{center}
\caption{Spectral-element grid for the VR grids in this study, (a) $\tt Arctic$, (b) $\tt Arctic-GrIS$ and (c) $\tt CONUS$. Note that this is not the computational grid; each element has $3\times3$ independent grid points.}
\label{fig:vr-grids}
\end{figure}

@@

The FV dycore is a hydrostatic model that integrates the equations of motion using a finite-volume discretization on a spherical lat-lon grid \cite{LR1997QJR}. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to an Eulerian reference grid in the vertical \cite{L2004MWR}, using a hybrid-pressure vertical coordinate. Hyperviscous damping is applied to the divergent modes, while Laplacian damping is applied to momentum in the top few layers, referred to as a \textit{sponge layer} \cite{L2011IJHPC} [phl: as far as I know the Laplacian-like damping is not invoked in lower resolution setups - was implemented for 1/4 applications where the polar night jet became excessively strong]. A polar filter damps computational instability due to the convergence of meridians, permitting a longer time step. It takes the form of a Fourier filter in the zonal direction, with the damping coefficients increasing monotonically in the meridional direction \cite{ST1995GEOS}. The form of the filter is designed to slow down the propagation of large zonal wave-numbers to satisfy the CFL condition of the shortest resolved wave at some reference latitude.

@@

The SE dycore supports regional grid refinement via its VR configuration, requiring two enhancements over uniform-resolution setups. Firstly, as the numerical viscosity increases with resolution, explicit hyperviscosity relaxes according to the local element size, reducing in strength  by an order of magnitude per halving of the grid spacing. A tensor-hyperviscosity formulation is used \cite{GetAl2014GMD}, which adjusts the coefficients in two orthogonal directions to more accurately target highly distorted quadrilateral elements. Secondly, the topography boundary conditions are smoothed in a way that does not excite grid scale modes, and so the NCAR topography software \cite{gmdd-8-4623-2015} has been modified to scale the smoothing radius by the local element size, resulting in rougher topography in the refinement zone.

For SE grids with quasi-uniform grid spacing, the SE tracer transport scheme is replaced with the Conservative Semi-Lagrangian Multi-tracer transport scheme (CSLAM) \cite<>{LTOUNGK2017MWR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a separate grid from the spectral-element dynamics, dividing each element into $3\times3$ control volumes with quasi-equal area. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the default SE dycore in which the physics are evaluated Gauss-Lobatto-Legendre points \cite{HL2018MWR}. {\color{purple}{Rene/Andrew - maybe elaborate further on the advantages of CSLAM?}}

@@

All grids and dycores in this study use 32 hybrid pressure-sigma levels in the vertical, with a model top of about 1~hPa or 40~km. However, note that any grid or dycore can in principle be run with a higher model top or finer vertical resolution.

@@

Initial simulations with the $\tt ne30pg3$ SE grid produced weaker shortwave cloud forcing relative to the tuned finite-volume dycore in the standard CESM2 configuration. The SE dycore in CESM2.2 therefore has two CLUBB parameter changes to provide more realistic cloud forcing and top-of-atmosphere radiation balance. We reduced the width of the sub-grid distribution of vertical velocity ($\tt clubb\_gamma=0.308\rightarrow0.270$) and also reduced the strength of the damping for horizontal component of turbulent energy ($\tt clubb\_c14=2.2\rightarrow1.6$) to increase cloudiness. For a description of how CLUBB parameters impact the simulated climate, see \citeA{GETAL2015JAMES}.

\subsection{Simulated surface mass balance (SMB)}

All grids and dycores simulate the GrIS Surface Mass Balance (SMB), which is the sum of mass accumulation of precipitation and mass loss from ablation. The latter is the sum of evaporation/sublimation and liquid runoff, with runoff being a combination of liquid precipitation and snow and ice melt. Not all liquid precipitation or snow/ice melt runs off the ice sheet; this water can penetrate pore spaces in the snowpack/firn layer and freeze, increasing the ice mass. These processes are represented by different CESM components, but it is the Community Land Model, version 5 \cite<CLM;>{CLM5}, that aggregates these processes and computes the SMB.

CLM runs on the same grid as the atmosphere, using a downscaling technique to account for sub-grid variability in SMB. In short, the ice sheet patch in a CLM grid cell is subdivided into 10 elevation classes (ECs), each with a distinct surface energy balance and SMB. The area fraction of each EC is derived from a high-resolution GrIS elevation dataset. The near-surface air temperature, humidity, and air density are calculated for each EC using an assumed lapse rate and the elevation difference from the grid-cell mean. Precipitation from CAM is repartitioned into solid or liquid based on the surface temperature of the EC; precipitation falls as snow for temperatures between $T < -2^\circ$~C, as rain for $T > 0^\circ$~C,  and as a linear combination of rain and snow for temperatures between $-2^\circ$~C and $0^\circ$~C.
Snow accumulation in each EC is limited to a depth of 10 m liquid water equivalent. Any snow above the 10-m cap is routed to the ocean as solid runoff.  Refreezing of liquid water within the snowpack is an additional source of ice.  Integrating over all ECs, weighting by the area fractions, provides a more accurate SMB than would be found using the grid-cell mean elevation. For a more detailed description of how the SMB is computed in CESM, we refer the reader to \citeA{CISM1,SETAL2019TC,KETAL2020JAMES,Muntjewerf_EA_2021_coupling}.

Changes in ice depth, but not snow depth, count toward the SMB.  That is, snow accumulation above the 10-m cap contributes a positive SMB, and surface ice melting (after melting of the overlying snow) yields a negative SMB.  Since snow in the accumulation zone must reach the cap to simulate a positive SMB, the snow depths on the VR grids were spun up by forcing CLM in standalone mode, cycling over data from a 20-year $\tt Arctic$ $\tt FHIST$ simulation (a model simulation with prescribed, observed sea-surface conditions) for about 500 years. The uniform-resolution grids are initialized with the SMB from an existing $\tt f09$ spun-up initial condition. In the simulations described here, the GrIS is prescribed at its observed, modern extent and thickness.

@@

We seek to integrate SMB components over a GrIS ice mask and to diagnose their contributions to the GrIS mass budget. However, the ice masks vary across the grids, especially in comparison to the RACMO3.2 ice mask, whose total area is about $3\%$ less than that of the reference dataset (Figure~\ref{fig:grisdx}). CLM's dataset creation tool generates the model ice mask by mapping a high-resolution dataset to the target grid using the Earth System Modeling Framework (ESMF) first-order conservative remapping algorithm \cite{ESMF}. The figure suggests that the mapping errors are less than $1.5\%$ across the CESM2 grids. The area errors in Figure~\ref{fig:grisdx} may  seem small, but even $1-2\%$ area differences can lead to large differences in integrated SMB \cite{HETAL2022TC}.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_grisres.pdf}
\end{center}
\caption{The spatial properties of the GrIS as represented by different grids in this study. (Left) GrIS area error, computed as the relative differences from a 4-km dataset used to create the CESM ice masks, (right) approximate average grid spacing over GrIS.}
\label{fig:grisdx}
\end{figure}

We have taken a common-ice-mask approach by mapping all model fields to the lowest-resolution grids, i.e., the $\tt f19$ and $\tt ne30pg2$ grids, and integrating over these low-resolution ice masks. The use of low-resolution common ice masks is a conservative decision, and is justified because we seek to use first-order remapping algorithms to map fields to the common ice mask, which is not generally reliable when mapping to a higher-resolution grid than the source grid. We use two remapping algorithms: ESMF first-order conservative and the TempestRemap \cite{TempestRemap} high-order monotone algorithm. Since mapping errors are sensitive to grid type, we evaluate all quantities on both common ice masks, the $\tt f19$ and $\tt ne30pg2$ masks. Thus, we evaluate an integrated quantity on a given grid up to four times to estimate the uncertainty due to differences in grid type and remapping algorithms.

The SMB is expressed in a form that is agnostic of water phase, a total water mass balance, to facilitate comparisons across different grids with different ice masks and to increase consistency with the variables available in the RACMO datasets. The SMB for total water can be expressed as:
\begin{equation}
SMB = accumulation + runoff + evaporation/sublimation, \label{eq:SMB}
\end{equation}
where all terms have consistent sign conventions (positive values contribute mass, and negative values reduce mass). The accumulation source term refers to the combined solid and liquid precipitation, runoff refers to the liquid water sink, and evaporation/sublimation is the vapor sink. Since the runoff term aggregates many processes, we isolate the melting contribution by also tracking the combined melt of snow and ice. Note that this SMB expression is different from the internally computed SMB in CLM \cite{KETAL2020JAMES}.

@@

Note that we are mapping to low-resolution grids that have larger GrIS areas than the source grids (Figure~\ref{fig:grisdx}). Since the components of equation~\ref{eq:SMB} are not confined to the ice mask, method~\ref{lbl:label1} reconstructs the SMB over the portion of the common ice mask that is outside the ice mask on the source grid. While this may be a an acceptable way to reconstruct the mass source terms over different ice masks, ice melt is zero outside the source ice mask, and so method~\ref{lbl:label1} will underestimate the mass sink term. This underestimation is systematic in method~\ref{lbl:label2}, where all variables are exclusive to the ice mask. Method~\ref{lbl:label2} is more dissipative, since mapping to the low-resolution grid will average a field of non-zero values over the ice mask with a field of zeros outside the ice mask. However, patch-level values for processes exclusive to the ice mask (e.g., ice melt) will on average have larger magnitudes than the the grid-mean quantities used in method~\ref{lbl:label1}. %The dilution of the these patch level fields caused by mapping to the common ice mask are then inflated due to integrating over the larger ice mask.
